Merging merges, more or less

### 2015-03-03 2015mergingmerges/section/01-Introduction.tex
rewrite intro

Our goal, then, appears relatively simple: to extend the regular dataflow languages enough to allow this subset of dynamic combinators, while keeping the desired properties, for the purpose of compilation and optimisation.

\subsection{Motivation}

Consider the following program, which partitions its input array by the first element of each pair.
If we plan on executing this program with data that is too large to execute in memory, what is the best execution strategy?

\begin{code}
app2 :: [(Bool,a)] -> [(Bool,a)]
app2 xs
  = filter     (fst) xs
 ++ filter (not.fst) xs

main
  = do xs  <- readMassiveFile
       sendEmails (app2 xs)
\end{code}

We may wish to iterate through the input array only once, but that would require buffering at least the second half of the append in memory, while the first half is being processed.
Laziness alone is insufficient here, as the second half of the append would still be referencing the start of the array as the first half is being pulled, leading to the same buffering.

Indeed, we believe that this case is in fact best solved by \emph{duplicating work}, to some extent:

\begin{code}
app2 :: [(Bool,a)] -> [(Bool,a)] -> [(Bool,a)]
app2 xs xs'
  = filter     (fst) xs
 ++ filter (not.fst) xs'

main
  = do xs  <- readMassiveFile
       xs' <- readMassiveFile
       sendEmails (app2 xs xs')
\end{code}

This is not always the case, however, and there are many situations where using the same input array multiple times does not require multiple traversals.

\begin{code}
zip2 :: [(Bool,a)] -> [(Bool,a)]
zip2 xs
  = map f xs `zip` map g xs

filter2 :: [(Bool,a)] -> [(Bool,a)]
filter2 xs
  = (filter f xs, filter g xs)
\end{code}
The goal of this work is to determine which programs can be executed with no buffering, and offer an efficient compilation strategy for these.

>>>>

In functional languages, fusion systems such as stream fusion\cite{coutts2007stream} and fold/build fusion\cite{jones2001playing}\TODO{correct?} rely on the inliner to move producers into their consumers, after which rewrite rules can remove intermediate allocations.
This short-cut fusion works well for vertical fusion when producers have only single consumers, but when a producer is used by multiple consumers it cannot be inlined without duplicating work, so fusion will not occur.
Our earlier work on flow fusion\cite{lippmeier2013data} is able to fuse these multiple consumer cases, but only for a small set of combinators; neither @append@ nor @merge@ are handled.

Our goal is to extend the regular dataflow languages enough to allow this subset of dynamic combinators, while keeping the desired properties, for the purpose of compilation and optimisation.

\subsection{Buffering}

Fusion for array programs is important for removing intermediate arrays, reducing memory traffic as well as allocations.
However, when dealing with data too large to fit in memory such as tables on disk, intermediate arrays becomes essential rather than just desirable.
Attempting to create an intermediate array of such amounts of data would lead to thrashing and swapping to disk, or perhaps even running out of swap.
For these situations, some sort of assurance of total fusion are required: either the program can be fused with no intermediate arrays or unbounded buffers, or it will not compile at all.

Consider the following program.
Suppose we have two sources of integers from disk, wish to filter each and pair them together, then perform some action for each pair:

\begin{code}
zip1 :: Source Int -> Source Int -> IO ()
zip1 xs ys
 = let xs'  = filter (>0) xs
       ys'  = filter (<0) ys
       zz   = zip xs' ys'
   in  mapM_ fire_missile zz
\end{code}

This program can be executed with no buffering of either input or any intermediate arrays.
Now suppose wish to perform another action over the first source:

\begin{code}
zip2 :: Source Int -> Source Int -> IO ()
zip2 xs ys
 = let xs'  = filter (>0) xs
       ys'  = filter (<0) ys
       xs2  = map    (+1) xs
       zz   = zip xs' ys'
   in  do   mapM_ fire_missile zz
            mapM_ fortify      xs2
\end{code}

Sadly, this program requires buffering the entire @xs@ source in memory while performing the first actions, then looping through the buffered @xs@ anf performing the second action afterwards.
However, we could execute this with no buffering if the relative ordering of the two @mapM_@s was not important, and could be reordered and interspersed as the data is read.
We use the following @When@ syntax to denote performing a reorderable IO action \emph{when}ever data is available.

\begin{code}
zip3 :: Source Int -> Source Int -> IO ()
zip3 xs ys
 = let xs'  = filter (>0) xs
       ys'  = filter (<0) ys
       xs2  = map    (+1) xs
       zz   = zip xs' ys'
   in  When fire_missile zz
   :&& When fortify      xs2
\end{code}

As a final example, suppose we removed the second data source and filtered the same source according to different predicates, then paired them together.
Can we still execute this with no buffer, and only iteration over @xs@?

\begin{code}
zip4 :: Source Int -> IO ()
zip4 xs
 = let xs'  = filter (>0) xs
       xs'' = filter (<0) xs
       zz   = zip xs' xs''
   in  mapM_ fire_missile zz
\end{code}

Sadly this last program requires an unbounded buffer or two iterations over @xs@, so we would like to disallow it.
It may not be immediately obvious just looking at the program that it requires two iterations.

In summary, the purpose of this work is to determine which programs can be executed with no buffering, and offer an efficient compilation strategy for these.
